---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r processing data,  echo = TRUE}
## Unzip activity.zip and extract its content into datafiles forlder
library(lubridate)
unzip(zipfile = "activity.zip", exdir = "./datafiles")
data <- as.data.frame (read.csv("./datafiles/activity.csv", header = TRUE))
data$date <- ymd(data$date)
```

## Histogram of the total number of steps taken each day

```{r sum of steps per day, ercho = TRUE}
library(dplyr)
library(xtable)
stepsxday <- data [!is.na(data$steps),] %>% group_by( by = date) %>%  summarize (sum(steps, na.rm=TRUE))
colnames(stepsxday)  <-  c("Day", "Steps")
xt  <- xtable(head(stepsxday))
print (xt, type="html")

```

```{r histogram, echo = TRUE}

hist(stepsxday$Steps,  main="steps per day",  ylab="Frecuency",  xlab="Steps",  border="red", col="green", las=1,breaks=10, xlim = c(0, 25000))

```


## What is mean total number of steps taken per day?


```{r mean of steps per day, ercho = TRUE}

summary(stepsxday$Steps)  

```
The mean is 10766 and de median is 10765 

## What is the average daily activity pattern?

```{r average spteps per 5 minutes interval, echo = "TRUE"}
stepsxinterval <- data [!is.na(data$steps),] %>% group_by( by = interval) %>%  summarize (mean(steps, na.rm=TRUE))
colnames(stepsxinterval)  <-  c("Interval", "Steps")
stepsxinterval 

```
```{r Time series plot of the average number of steps taken, echo= TRUE, fig.width=10}
library(ggplot2)
ggplot(data =stepsxinterval, aes(x=Interval, y= Steps)) + geom_line()
```

## The 5-minute interval that, on average, contains the maximum number of steps

To identify the exact five minute slot with the higher average of steps I only have to order the results from the previous calculation

```{r Identify the interval whit the gratest mean os steps, echo = TRUE}
 head(stepsxinterval[order(-stepsxinterval$Steps),])
```
The 5-minute interval that, on average, contains the maximum number of steps is the 835

## Imputing missing values

I will replace each NA five minutes intervals by its average

```{r Imputing missing values, echo =TRUE}
dataMod <- data
dataMod$interval <-as.factor(dataMod$interval)
dataMod$StepsMeanInt<- tapply(dataMod$steps, dataMod$interval, mean, na.rm = TRUE)
dataMod <- mutate(dataMod, stepsMod = ifelse(is.na(dataMod$steps), dataMod$StepsMeanInt, dataMod$steps))
```

Next I will calculate the sum of steps by date using the new steps column with  imputed missing data

```{r sum of steps per day (without missing values), ercho = TRUE}
stepsxdayMod <- data [!is.na(data$steps),] %>% group_by( by = date) %>%  summarize (sum(steps, na.rm=TRUE))
colnames(stepsxdayMod)  <-  c("Day", "Steps")
xt  <- xtable(head(stepsxdayMod))
print (xt, type="html")
```

```{r histogram without missing values, echo = TRUE}
hist(stepsxdayMod$Steps,  main="steps per day (missing data as an average) ",  ylab="Frecuency",  xlab="Steps",  border="red", col="green", las=1,breaks=10, xlim = c(0, 25000))
```


## Are there differences in activity patterns between weekdays and weekends?

```{r weekday /weekend differentiation, echo = TRUE}
dataMod <- mutate (dataMod[!is.na(dataMod$steps),  ], WeekPeriod  = ifelse(wday( date, week_start = 1)< 6 , "Weekday" , "Wekend") )
dataMod <- select (dataMod, !4:5)
head(dataMod)
```
```{r average spteps per 5 minutes intervaland day of the week, echo = "TRUE"}
dataMod$WeekPeriod <- as.factor (dataMod$WeekPeriod )
IntW  <- dataMod %>% group_by( interval, WeekPeriod )%>% summarise(Ninter = mean(steps))
head(IntW)
```
```{r Time series plot of the average number of steps taken by weekday and weekend, echo= TRUE, fig.width= 10}
g <- ggplot (data = IntW, aes(x= IntW$interval, y = IntW$Ninter))
g  + geom_point () + geom_line (aes( group=1))+ facet_grid(~ WeekPeriod) + 
  xlab("5 minute interval") +  ylab ("Average of steps")

```
During weekdays, a lot of walking is at the beginning of the day and during the weekends is more spread during the day.
